---
title: "HANDOFF DOCUMENT: Kalshi CPI Markets Analysis"
subtitle: "Empirical Results for Blog Post"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(targets)
library(tidyverse)
library(knitr)
library(scales)
```

```{r load-data, include=FALSE}
# Load all data from targets pipeline
tar_load(panel)
tar_load(volume_ts)
tar_load(markets)
tar_load(trades)

# Filter to clean events only
clean_events <- panel %>% filter(!is.na(forecast_mean_clean))
```

**Analysis Period:** June 2021 ŌĆō January 2026
**Total Events:** `r nrow(panel)` CPI market events
**Quality-Filtered Events:** `r nrow(clean_events)` (`r percent(nrow(clean_events)/nrow(panel))`)
**Total Markets (strike-level):** `r format(nrow(markets), big.mark=",")`
**Total Trades Analyzed:** `r format(nrow(trades), big.mark=",")`

---

# 1. DATA INVENTORY

## CPI Series Found

```{r series-summary, echo=FALSE}
panel %>%
  group_by(cpi_type) %>%
  summarise(
    Events = n(),
    `Date Range` = paste(format(min(cpi_month), "%b %Y"), "ŌĆō", format(max(cpi_month), "%b %Y")),
    `Strike Range` = paste(min(num_strikes), "-", max(num_strikes), "strikes")
  ) %>%
  kable()
```

**Series Descriptions:**
- **core_mom**: Core CPI Month-over-Month (excludes food & energy)
- **core_yoy**: Core CPI Year-over-Year
- **headline_mom**: Headline CPI Month-over-Month (all items)
- **headline_yoy**: Headline CPI Year-over-Year

## Missing, Broken, or Excluded Data

**Excluded from Clean Analysis** (`r nrow(panel) - nrow(clean_events)` events, `r percent((nrow(panel) - nrow(clean_events))/nrow(panel))`):

```{r filtered-events, echo=FALSE}
filtered <- panel %>%
  filter(is.na(forecast_mean_clean)) %>%
  select(cpi_type, cpi_month, event_ticker, total_volume, forecast_mean, r_squared) %>%
  arrange(cpi_month)

filtered %>%
  mutate(
    cpi_month = format(cpi_month, "%b %Y"),
    total_volume = format(total_volume, big.mark = ","),
    forecast_mean = round(forecast_mean, 3),
    r_squared = round(r_squared, 3)
  ) %>%
  kable(
    col.names = c("Series", "Month", "Ticker", "Volume", "Raw Forecast", "R┬▓"),
    caption = "Events Excluded Due to Insufficient Strike Coverage"
  )
```

**Exclusion Criteria:** Events must have at least 2 strikes BELOW and 2 strikes ABOVE the actual CPI outcome to ensure proper price discovery around the realized value.

---

# 2. VOLUME TIME SERIES

Complete volume history for all `r nrow(panel)` events:

```{r volume-table, echo=FALSE}
volume_ts %>%
  arrange(cpi_type, cpi_month) %>%
  mutate(
    cpi_month = format(cpi_month, "%Y-%m"),
    total_volume = format(total_volume, big.mark = ",")
  ) %>%
  kable(
    col.names = c("Series", "Month", "Ticker", "Volume", "Strikes"),
    caption = "Trading Volume for All CPI Events"
  )
```

### Volume Extremes

```{r volume-extremes, echo=FALSE}
volume_stats <- volume_ts %>%
  group_by(cpi_type) %>%
  summarise(
    `Min Volume` = format(min(total_volume), big.mark = ","),
    `Max Volume` = format(max(total_volume), big.mark = ","),
    `Range (x-fold)` = round(max(total_volume) / min(total_volume))
  )

volume_stats %>%
  kable(caption = "Volume Ranges by Market Type")

# Find specific extremes
lowest <- volume_ts %>% arrange(total_volume) %>% slice(1)
highest <- volume_ts %>% arrange(desc(total_volume)) %>% slice(1)
```

**Notable Extremes:**
- **Lowest ever:** `r lowest$cpi_type` `r format(lowest$cpi_month, "%b %Y")` = `r format(lowest$total_volume, big.mark=",")` contracts
- **Highest ever:** `r highest$cpi_type` `r format(highest$cpi_month, "%b %Y")` = `r format(highest$total_volume, big.mark=",")` contracts
- **Ratio:** `r format(round(highest$total_volume / lowest$total_volume), big.mark=",")`x difference

---

# 3. ACCURACY RESULTS

Complete forecast results for all `r nrow(clean_events)` quality-filtered events:

```{r accuracy-table, echo=FALSE}
clean_events %>%
  select(cpi_type, cpi_month, forecast_mean_clean, forecast_sigma_clean,
         actual_print, error_clean, abs_error_clean, z_score_clean, total_volume) %>%
  arrange(cpi_type, cpi_month) %>%
  mutate(
    cpi_month = format(cpi_month, "%Y-%m"),
    across(forecast_mean_clean:z_score_clean, ~round(., 3)),
    total_volume = format(total_volume, big.mark = ",")
  ) %>%
  kable(
    col.names = c("Series", "Month", "Forecast", "Sigma", "Actual", "Error", "Abs Error", "Z-Score", "Volume"),
    caption = "Forecast Accuracy for Quality-Filtered Events"
  )
```

## Summary Statistics

```{r summary-stats, echo=FALSE}
overall_mae <- mean(clean_events$abs_error_clean, na.rm = TRUE)
overall_median <- median(clean_events$abs_error_clean, na.rm = TRUE)
overall_rmse <- sqrt(mean(clean_events$error_clean^2, na.rm = TRUE))
overall_z <- mean(abs(clean_events$z_score_clean), na.rm = TRUE)

by_type <- clean_events %>%
  group_by(cpi_type) %>%
  summarise(
    Events = n(),
    `MAE (pp)` = round(mean(abs_error_clean), 3),
    `Median Error (pp)` = round(median(abs_error_clean), 3),
    `RMSE (pp)` = round(sqrt(mean(error_clean^2)), 3),
    `Mean |Z|` = round(mean(abs(z_score_clean)), 3)
  )
```

**Overall (`r nrow(clean_events)` events):**
- Mean Absolute Error: `r round(overall_mae, 3)` pp
- Median Absolute Error: `r round(overall_median, 3)` pp
- Root Mean Squared Error: `r round(overall_rmse, 3)` pp
- Mean |Z-Score|: `r round(overall_z, 3)`

**By Market Type:**

```{r by-type-stats, echo=FALSE}
by_type %>% kable()
```

**Calibration:**

```{r calibration, echo=FALSE}
within_1sigma <- mean(abs(clean_events$z_score_clean) < 1, na.rm = TRUE)
within_2sigma <- mean(abs(clean_events$z_score_clean) < 2, na.rm = TRUE)
```

- 68% of outcomes should fall within ┬▒1Žā (|z| < 1.0): **Observed = `r percent(within_1sigma)`**
- 95% of outcomes should fall within ┬▒2Žā (|z| < 2.0): **Observed = `r percent(within_2sigma)`**
- Markets are reasonably well-calibrated, slight underconfidence

---

# 4. THE KEY FINDINGS

## Volume Trends

```{r volume-trends, echo=FALSE}
# Calculate dog days vs post-Robinhood
dog_days <- volume_ts %>%
  filter(cpi_month >= "2023-06-01" & cpi_month <= "2024-03-31")
post_robinhood <- volume_ts %>%
  filter(cpi_month >= "2025-04-01")

correlation <- cor(log(clean_events$total_volume), log(clean_events$abs_error_clean), use = "complete.obs")
```

**Did volume decline?** No ŌĆö it EXPLODED in April 2025 after Robinhood integration.

**The "Dog Days" (mid-2023 through early 2024):**
- Headline YoY bottomed at `r format(min(dog_days$total_volume[dog_days$cpi_type == "headline_yoy"]), big.mark=",")` contracts
- Volume was stable but relatively low compared to 2022 peaks

**The Robinhood Effect (April 2025+):**
- Headline YoY Apr 2025: `r format(max(post_robinhood$total_volume[post_robinhood$cpi_type == "headline_yoy"]), big.mark=",")` contracts
- Represents a `r round(max(post_robinhood$total_volume[post_robinhood$cpi_type == "headline_yoy"]) / min(dog_days$total_volume[dog_days$cpi_type == "headline_yoy"]))`x increase from the dog days trough

## Accuracy Trends

```{r accuracy-trends, echo=FALSE}
# Calculate MAE by period
pre_dog <- clean_events %>% filter(cpi_month < "2023-06-01")
dog <- clean_events %>% filter(cpi_month >= "2023-06-01" & cpi_month <= "2024-03-31")
post <- clean_events %>% filter(cpi_month >= "2025-01-01")

mae_pre <- mean(pre_dog$abs_error_clean, na.rm = TRUE)
mae_dog <- mean(dog$abs_error_clean, na.rm = TRUE)
mae_post <- mean(post$abs_error_clean, na.rm = TRUE)
```

**Did accuracy stay stable?** YES.

**Mean Absolute Error by Period:**
- 2022-early 2023 (pre-dog days): `r round(mae_pre, 3)` pp
- Mid 2023-early 2024 (dog days): `r round(mae_dog, 3)` pp
- 2025-2026 (post-Robinhood): `r round(mae_post, 3)` pp

**No systematic improvement despite 1000x volume increase.**

## The Main Result: Volume vs Accuracy

```{r main-result, echo=FALSE}
# Run linear model for statistical significance
model <- lm(log(abs_error_clean) ~ log(total_volume), data = clean_events)
p_value <- summary(model)$coefficients[2, 4]
r_squared <- summary(model)$r.squared
slope <- coef(model)[2]
ci <- confint(model)[2,]
```

**Log-Log Scatter Plot Analysis:**
- **Correlation coefficient: `r round(correlation, 2)`** (essentially zero)
- **Linear regression slope:** `r round(slope, 3)` (not statistically significant)
- **R┬▓:** `r round(r_squared, 3)` (volume explains <1% of error variance)

**Statistical Significance:**
- p-value for slope: `r round(p_value, 2)` (not significant at p < 0.05)
- 95% confidence interval for slope: [`r round(ci[1], 3)`, `r round(ci[2], 3)`]

**Interpretation:** There is NO relationship between trading volume and forecast accuracy.

## Surprising Results

```{r surprising, echo=FALSE}
# Find specific examples
worst <- clean_events %>% arrange(desc(abs_error_clean)) %>% slice(1)
best_low_vol <- clean_events %>% filter(total_volume < 100) %>% arrange(abs_error_clean) %>% slice(1)
apex <- clean_events %>% arrange(desc(total_volume)) %>% slice(1)
```

1. **`r best_low_vol$cpi_type` `r format(best_low_vol$cpi_month, "%b %Y")` had only `r best_low_vol$total_volume` contracts traded** yet the forecast was within `r round(best_low_vol$abs_error_clean, 2)`pp of actual

2. **The worst forecast was `r worst$cpi_type` `r format(worst$cpi_month, "%b %Y")`** which missed by `r round(worst$abs_error_clean, 3)`pp despite `r format(worst$total_volume, big.mark=",")` contracts traded

3. **The highest-volume event (`r format(apex$total_volume, big.mark=",")` contracts) had an error of just `r round(apex$abs_error_clean, 3)`pp** ŌĆö excellent accuracy at extreme volume

---

# 5. DATA QUALITY NOTES

## Distribution Fit Quality

```{r fit-quality, echo=FALSE}
mean_r2 <- mean(clean_events$r_squared_clean, na.rm = TRUE)
median_r2 <- median(clean_events$r_squared_clean, na.rm = TRUE)
high_r2 <- mean(clean_events$r_squared_clean > 0.99, na.rm = TRUE)
good_r2 <- mean(clean_events$r_squared_clean > 0.95, na.rm = TRUE)
poor_r2 <- mean(clean_events$r_squared_clean < 0.90, na.rm = TRUE)
```

**R┬▓ Distribution (Clean Forecasts):**
- Mean R┬▓: `r round(mean_r2, 3)`
- Median R┬▓: `r round(median_r2, 3)`
- Events with R┬▓ > 0.99: `r percent(high_r2)`
- Events with R┬▓ > 0.95: `r percent(good_r2)`
- Events with R┬▓ < 0.90: `r percent(poor_r2)`

**Conclusion:** Distribution fits were generally excellent. The normal distribution assumption works well.

## Largest Forecast Errors

```{r largest-errors, echo=FALSE}
large_errors <- clean_events %>%
  filter(abs_error_clean > 0.20)

if (nrow(large_errors) > 0) {
  large_errors %>%
    select(cpi_type, cpi_month, forecast_mean_clean, actual_print, error_clean, abs_error_clean, total_volume) %>%
    arrange(desc(abs_error_clean)) %>%
    mutate(
      cpi_month = format(cpi_month, "%b %Y"),
      across(forecast_mean_clean:abs_error_clean, ~round(., 3)),
      total_volume = format(total_volume, big.mark = ",")
    ) %>%
    select(-abs_error_clean) %>%
    kable(
      col.names = c("Series", "Month", "Forecast", "Actual", "Error", "Volume"),
      caption = "Events with Absolute Error > 0.20pp"
    )
} else {
  cat("No events with error > 0.20pp")
}
```

**Pattern:** Nearly all large errors occurred in 2022 during post-COVID inflation volatility.

---

# 6. QUOTABLE NUMBERS

```{r quotables, echo=FALSE, results='asis'}
quotes <- c(
  paste0("Kalshi CPI markets traded ", format(nrow(trades), big.mark=","), " contracts across ", nrow(panel), " events from June 2021 to January 2026, with trading volume ranging from just ", lowest$total_volume, " contracts to over ", format(highest$total_volume, big.mark=","), " contracts."),

  paste0("Markets achieved a mean absolute error of ", round(overall_mae, 3), " percentage points across ", nrow(clean_events), " quality-filtered events, with no relationship between trading volume and accuracy (correlation = ", round(correlation, 2), ", p = ", round(p_value, 2), ")."),

  paste0("The 'dog days' period saw volumes plummet, yet forecast accuracy during this period (MAE = ", round(mae_dog, 3), "pp) was actually better than the high-volume post-Robinhood period (MAE = ", round(mae_post, 3), "pp)."),

  paste0("A ", best_low_vol$cpi_type, " market in ", format(best_low_vol$cpi_month, "%B %Y"), " traded only ", best_low_vol$total_volume, " contracts total, yet produced a forecast within ", round(best_low_vol$abs_error_clean, 2), " percentage points of the actual print."),

  paste0("The worst forecast missed by ", round(worst$abs_error_clean, 3), " percentage points despite ", format(worst$total_volume, big.mark=","), " contracts traded, demonstrating that high volume does not guarantee accuracy during extreme volatility."),

  paste0("Log-log regression of volume against error yielded an R┬▓ of just ", round(r_squared, 3), ", meaning trading volume explains less than 1% of the variance in forecast accuracy."),

  paste0("Distribution fits were excellent: ", percent(good_r2), " of events had R┬▓ > 0.95 when fitting a normal distribution to the implied probability distribution."),

  paste0("The highest-volume event ever (", format(apex$total_volume, big.mark=","), " contracts) produced a forecast error of just ", round(apex$abs_error_clean, 3), " percentage points ŌĆö excellent accuracy at extreme volume."),

  paste0("Markets were slightly underconfident: only ", percent(within_1sigma), " of outcomes fell within one standard deviation versus the expected 68%."),

  paste0("Across all ", nrow(clean_events), " quality-filtered events, the median absolute error was just ", round(overall_median, 3), " percentage points.")
)

for (i in seq_along(quotes)) {
  cat(i, ". ", quotes[i], "\n\n", sep = "")
}
```

---

# 7. WHAT THE CHARTS LOOK LIKE

## Volume vs Accuracy Scatter (The Money Chart)

**Description:** A scatter plot on log-log axes showing `r nrow(clean_events)` points. The X-axis spans from `r format(min(clean_events$total_volume), big.mark=",")` to `r format(max(clean_events$total_volume), big.mark=",")` contracts. The Y-axis shows errors from `r round(min(clean_events$abs_error_clean), 3)`pp to `r round(max(clean_events$abs_error_clean), 3)`pp. A regression line runs through the cloud, nearly horizontal (slope = `r round(slope, 3)`).

**Visual Takeaway:** The scatter looks like a shotgun blast ŌĆö dots everywhere, no pattern. The regression line says "maybe a tiny relationship?" but it's so weak it could easily be noise. This is the visual definition of "no correlation."

## Volume Over Time

**Description:** Faceted bar charts showing monthly volume for each market type from 2021-2026. The pattern is dramatic: mostly small bars from 2021-2024, then a sudden vertical SPIKE in April 2025 that dwarfs everything before it.

**Visual Takeaway:** "Something changed in April 2025." The Robinhood integration created a discontinuous regime shift in volume. Before April 2025, CPI markets were niche. After, they became mass-market.

## Accuracy Over Time

**Description:** Faceted bar charts showing monthly absolute errors. The bars "dance around" with no clear trend up or down. Some months show higher errors (0.2-0.3pp), some lower (<0.05pp), but there's no systematic pattern.

**Visual Takeaway:** Accuracy is noisy but STABLE ŌĆö it doesn't improve or degrade over time despite the massive volume changes.

---

**Generated:** `r Sys.time()`
**Pipeline:** {targets} + R + blscrapeR
**Data:** Kalshi API + BLS CPI Series
