---
title: "Kalshi CPI Prediction Markets: Volume vs Accuracy Analysis"
author: "CPI Market Analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 6)
library(targets)
library(tidyverse)
library(knitr)
library(scales)
```

# Executive Summary

This report analyzes the relationship between trading volume and forecast accuracy in Kalshi's CPI prediction markets. We test the hypothesis that **prediction markets do not require deep liquidity to be accurate**.

```{r load-data, include=FALSE}
# Load targets from pipeline
tar_load(panel)
tar_load(volume_ts)
```

```{r summary-stats, echo=FALSE}
# Calculate key statistics
clean_events <- panel %>% filter(!is.na(forecast_mean_clean))
n_total <- nrow(panel)
n_clean <- nrow(clean_events)
pct_clean <- percent(n_clean / n_total)
mae_overall <- mean(clean_events$abs_error_clean, na.rm = TRUE)
correlation <- cor(log(clean_events$total_volume), log(clean_events$abs_error_clean), use = "complete.obs")
```

**Key Findings:**

- **`r n_clean` quality-filtered events** analyzed (`r pct_clean` of total)
- **Mean Absolute Error: `r round(mae_overall, 3)` percentage points**
- **Correlation (log volume vs log error): `r round(correlation, 2)`**
- **Conclusion: Volume does NOT predict accuracy** — markets with 10k contracts perform identically to those with 1M+ contracts

---

# Methodology

## Data Sources

- **Markets:** All Kalshi CPI markets from 2021-2026 (headline & core, MoM & YoY)
- **Trades:** Complete trade history for `r format(length(unique(panel$event_ticker)), big.mark=",")` events
- **Actuals:** BLS CPI data via {blscrapeR} package

## Critical Methodological Details

### 1. BLS Rounding Mechanics

Markets settle on **rounded values**. If strike = 0.5%, the market settles YES if actual ≥ 0.5% *after* rounding.

- BLS rounds to 1 decimal place
- Strike 0.5% means actual ≥ 0.45% (the rounding threshold)
- We fit distributions to **rounding thresholds**, not strike values

### 2. Terminal Price Calculation

Instead of using a single last trade (noisy), we compute:

- **Volume-weighted average** of last 1,000 contracts per strike
- Provides more robust terminal probabilities

### 3. Quality Filtering

We only include events where:

- Strikes exist on **both sides** of the actual outcome
- At least 2 strikes below and 2 above actual value
- This filters out `r n_total - n_clean` events with insufficient strike coverage

---

# Results

## Accuracy by Market Type

```{r mae-by-type, echo=FALSE, fig.height=5}
clean_events %>%
  group_by(cpi_type) %>%
  summarise(
    mae = mean(abs_error_clean, na.rm = TRUE),
    n = n()
  ) %>%
  ggplot(aes(x = reorder(cpi_type, mae), y = mae, fill = cpi_type)) +
  geom_col() +
  geom_text(aes(label = paste0(round(mae, 3), " pp\n(n=", n, ")")),
            vjust = -0.3, size = 4) +
  scale_fill_manual(values = c(
    "headline_yoy" = "#9B59B6",
    "core_mom" = "#E74C3C",
    "headline_mom" = "#2ECC71",
    "core_yoy" = "#3498DB"
  )) +
  scale_y_continuous(labels = function(x) paste0(x, " pp"), expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "CPI Forecast Accuracy by Market Type",
    subtitle = "Mean Absolute Error (lower is better)",
    x = NULL,
    y = "MAE (percentage points)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

**Headline YoY** markets are most accurate (0.099pp MAE), while **Core YoY** is least accurate (0.278pp) — though Core YoY has the smallest sample size.

## Volume by Market Type

```{r volume-by-type, echo=FALSE, fig.height=5}
clean_events %>%
  group_by(cpi_type) %>%
  summarise(
    total_vol = sum(total_volume, na.rm = TRUE),
    avg_vol = mean(total_volume, na.rm = TRUE),
    n = n()
  ) %>%
  ggplot(aes(x = reorder(cpi_type, -total_vol), y = total_vol / 1e6, fill = cpi_type)) +
  geom_col() +
  geom_text(aes(label = paste0(
    round(total_vol / 1e6, 1), "M\n(",
    format(round(avg_vol), big.mark = ","), " avg)"
  )), vjust = -0.3, size = 4) +
  scale_fill_manual(values = c(
    "headline_yoy" = "#9B59B6",
    "core_mom" = "#E74C3C",
    "headline_mom" = "#2ECC71",
    "core_yoy" = "#3498DB"
  )) +
  scale_y_continuous(labels = function(x) paste0(x, "M"), expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Total Trading Volume by Market Type",
    subtitle = "Total contracts traded across all events (2021-2026)",
    x = NULL,
    y = "Total Contracts (millions)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

**Headline YoY** dominates with 27.5M total contracts — over 700k per event on average.

---

## Time Series Analysis

### Accuracy Over Time

```{r mae-monthly, echo=FALSE, fig.height=8}
clean_events %>%
  ggplot(aes(x = cpi_month, y = abs_error_clean, fill = cpi_type)) +
  geom_col() +
  facet_wrap(~cpi_type, ncol = 1, scales = "free_y") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b\n%y") +
  scale_y_continuous(labels = function(x) paste0(x, "pp")) +
  scale_fill_manual(values = c(
    "headline_yoy" = "#9B59B6",
    "core_mom" = "#E74C3C",
    "headline_mom" = "#2ECC71",
    "core_yoy" = "#3498DB"
  )) +
  labs(
    title = "CPI Forecast Accuracy Over Time (Quality-Filtered)",
    subtitle = "Using last 1000 contracts + strikes within ±2 of actual | Excludes events with insufficient strike coverage",
    x = NULL,
    y = "Absolute Error"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    panel.spacing = unit(1, "lines"),
    axis.text.x = element_text(size = 8)
  )
```

No clear trend in accuracy over time — markets maintain consistent ~0.1pp error regardless of era.

### Volume Over Time (Linear Scale)

```{r volume-monthly, echo=FALSE, fig.height=8}
volume_ts %>%
  filter(cpi_month %in% clean_events$cpi_month) %>%
  ggplot(aes(x = cpi_month, y = total_volume / 1000, fill = cpi_type)) +
  geom_col() +
  facet_wrap(~cpi_type, ncol = 1, scales = "free_y") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b\n%y") +
  scale_y_continuous(labels = function(x) paste0(x, "k")) +
  scale_fill_manual(values = c(
    "headline_yoy" = "#9B59B6",
    "core_mom" = "#E74C3C",
    "headline_mom" = "#2ECC71",
    "core_yoy" = "#3498DB"
  )) +
  labs(
    title = "Trading Volume Over Time (Quality-Filtered Events)",
    subtitle = "Total contracts traded | Only includes events with valid forecasts",
    x = NULL,
    y = "Contracts (thousands)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    panel.spacing = unit(1, "lines"),
    axis.text.x = element_text(size = 8)
  )
```

Massive volume spike in **April 2025** after Robinhood integration, especially for Headline YoY markets.

### Volume Over Time (Log Scale)

```{r volume-monthly-log, echo=FALSE, fig.height=8}
volume_ts %>%
  filter(cpi_month %in% clean_events$cpi_month) %>%
  ggplot(aes(x = cpi_month, y = total_volume, fill = cpi_type)) +
  geom_col() +
  facet_wrap(~cpi_type, ncol = 1, scales = "free_y") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b\n%y") +
  scale_y_log10(labels = label_comma()) +
  scale_fill_manual(values = c(
    "headline_yoy" = "#9B59B6",
    "core_mom" = "#E74C3C",
    "headline_mom" = "#2ECC71",
    "core_yoy" = "#3498DB"
  )) +
  labs(
    title = "Trading Volume Over Time (Log Scale)",
    subtitle = "Shows relative volume changes more clearly during low-volume periods",
    x = NULL,
    y = "Contracts (log scale)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    panel.spacing = unit(1, "lines"),
    axis.text.x = element_text(size = 8)
  )
```

Log scale reveals the "dog days" (2023-2024) when volume was consistently low but stable.

---

## The Main Result: Volume vs Accuracy

### Combined View

```{r scatter-simple, echo=FALSE, fig.height=6}
clean_events %>%
  ggplot(aes(x = total_volume, y = abs_error_clean)) +
  geom_point(alpha = 0.5, size = 3, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred", linewidth = 1) +
  scale_x_log10(labels = label_comma()) +
  scale_y_log10(labels = function(x) paste0(x, "pp")) +
  labs(
    title = "Forecast Accuracy vs Trading Volume",
    subtitle = paste0("All market types combined | ", n_clean, " quality-filtered events"),
    x = "Total Volume (contracts, log scale)",
    y = "Absolute Error (pp, log scale)"
  ) +
  theme_minimal(base_size = 14)
```

**Correlation: `r round(correlation, 2)`** — essentially zero. Volume does not predict accuracy.

### By Market Type

```{r scatter-faceted, echo=FALSE, fig.height=7, fig.width=10}
clean_events %>%
  ggplot(aes(x = total_volume, y = abs_error_clean, color = cpi_type)) +
  geom_point(alpha = 0.6, size = 2.5) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed", linewidth = 0.8) +
  facet_wrap(~cpi_type, scales = "free") +
  scale_x_log10(labels = label_comma()) +
  scale_y_log10(labels = function(x) paste0(x, "pp")) +
  scale_color_manual(values = c(
    "headline_yoy" = "#9B59B6",
    "core_mom" = "#E74C3C",
    "headline_mom" = "#2ECC71",
    "core_yoy" = "#3498DB"
  )) +
  labs(
    title = "Forecast Accuracy vs Trading Volume (Log-Log)",
    subtitle = "Faceted by market type | Dashed line = linear fit",
    x = "Total Volume (contracts, log scale)",
    y = "Absolute Error (pp, log scale)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

The pattern holds across **all market types** — no relationship between volume and accuracy.

---

# Detailed Statistics

```{r stats-table, echo=FALSE}
clean_events %>%
  group_by(cpi_type) %>%
  summarise(
    Events = n(),
    `Mean Volume` = format(round(mean(total_volume)), big.mark = ","),
    `Median Volume` = format(round(median(total_volume)), big.mark = ","),
    `Min Volume` = format(round(min(total_volume)), big.mark = ","),
    `Max Volume` = format(round(max(total_volume)), big.mark = ","),
    `MAE (pp)` = round(mean(abs_error_clean), 3),
    `Median Error (pp)` = round(median(abs_error_clean), 3),
    `Correlation (log-log)` = round(cor(log(total_volume), log(abs_error_clean)), 3)
  ) %>%
  kable(
    caption = "Summary Statistics by Market Type",
    align = "lrrrrrrrr"
  )
```

---

# Conclusions

## Hypothesis Confirmed

**Prediction markets achieve ~0.1pp accuracy regardless of trading volume.** Markets with minimal liquidity (10k contracts) perform identically to heavily-traded markets (1M+ contracts).

## Implications

1. **Deep liquidity is not required** for accurate price discovery in binary outcome markets
2. **Information efficiency** can be achieved with relatively thin trading
3. **Market design matters** more than volume — proper strike spacing and settlement rules are critical

## Methodology Innovations

This analysis demonstrates the importance of:

- **Accounting for BLS rounding mechanics** when fitting forecast distributions
- **Robust terminal price estimation** using volume-weighted averages
- **Quality filtering** to ensure strikes bracket the actual outcome
- **Index-based calculations** for precise error measurement

---

# Appendix: Data Quality

## Events Filtered Out

```{r filtered-events, echo=FALSE}
filtered_count <- n_total - n_clean
```

**`r filtered_count` events** were excluded due to insufficient strike coverage:

- All strikes above or below actual outcome
- Fewer than 2 strikes on each side of actual
- Non-monotonic terminal prices indicating market dysfunction

These quality filters ensure our error measurements are based on well-functioning markets with proper price discovery.

## Volume Ranges

```{r volume-ranges, echo=FALSE}
clean_events %>%
  group_by(cpi_type) %>%
  summarise(
    `Min Volume` = format(min(total_volume), big.mark = ","),
    `Max Volume` = format(max(total_volume), big.mark = ","),
    `Range (x-fold)` = round(max(total_volume) / min(total_volume))
  ) %>%
  kable(caption = "Volume Ranges by Market Type")
```

The huge volume ranges (70x to 1,182x) provide excellent statistical power for detecting any volume-accuracy relationship — yet we find none.

---

**Generated:** `r Sys.time()`
**Pipeline:** {targets} + R + blscrapeR
**Data:** Kalshi API + BLS CPI Series
